---
title: "How to Make Heatmap"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Heatmap is widely used method for visualization of geneexpression data. I make heatmap for visualization of log2FC data from multiple samples. This time I decided to write a notebook by breaking down every steps so that it become easier for me too modify this code to produce desired heatmap.

First we need to load all the package we are going to use. 

```{r}
library(ComplexHeatmap)
library(circlize)
```

Load gene expression data for clustering.

```{r}
expr = readRDS(system.file(package = "ComplexHeatmap", "extdata", "gene_expression.rds"))
```
Select column with expression value and convert them to matrix because data need to be in matrix formate to use as input for Heatmap function.

```{r}
mat = as.matrix(expr[, grep("cell", colnames(expr))])
```

In order to get nice clustering we need to scale data using scale function of R. Scale function center the data and convert it to z-score. In code bellow we apply scale function along the column of the matrix.

```{r}
mat_scaled = t(apply(mat, 1, scale))
```
We generated a vector of row means to annotate the heatmap with raw mean value of each genes.
```{r}
base_mean = rowMeans(mat)
```

Here we are using regular expression to extract cell type from the column names.
```{r}
type = gsub("s\\d+_", "", colnames(mat))
```
Next we are using cell types vector to generate a HeatmapAnnotation object to annotate the column of of the heatmap.
```{r}
ha = HeatmapAnnotation(type = type, annotation_name_side = "left")
```

Now let gradually build the complext heatmap. At first lets just took at the heatmap of our scalled expression data.
```{r}
Heatmap(mat_scaled, name = "expression")
```
Now lets change the color of the heatmap adding values to col option, here we give two vectors as input one vector for break value of our gene expression data and another vector that correspond to the break value.
```{r}
colors = colorRamp2(c(-2, 0, 2), c("green", "white", "red"))
Heatmap(mat_scaled, name = "expression", col = colors)
```
In the above heatmap we can not see the row name clearly so I would like to remove the reowname by setting show_row_names to FALSE.

```{R}
Heatmap(mat_scaled, name = "Expression", col = colors, row_title = NULL, show_row_dend = FALSE, show_row_names = FALSE)
```
Now we would like to add some annotation to the heatmap, let add annotation for column in the heatmap. In this heatmap columns are cells, there are three types of cells in this data. A the type vector that we defined earilier is the vector of cell type for each column. The length of vector is equal to the number of column. With the code below we will make a HeatmapAnnotation object ha then use it to annotate the column of heatmap. We decided to put column annotation on the top of the heatmap using top_annotation option. 

```{r}
ha = HeatmapAnnotation(type = type, annotation_name_side = "left")
Heatmap(mat_scaled, name = "Expression", col = colors, row_title = NULL, show_row_dend = FALSE, show_row_names = FALSE, top_annotation = ha)
```
We can see that three color bar at the top of the heatmap represent three different cell type and ligend for the bar color at the left side of the heatmap.

Now lets break the heatmap into five different clusters, by setting us row_km to 5. 

```{r}
Heatmap(mat_scaled, name = "Expression", col = colors, row_title = NULL, show_row_dend = FALSE, show_row_names = FALSE, top_annotation = ha, row_km = 5)
```

We can add another heatmap that represent mean value of gene's expression acrros the row of our data. Earlier we made a vector with the mean value of the row. 

```{r}
Heatmap(mat_scaled, name = "Expression", col = colors, row_title = NULL, show_row_dend = FALSE, show_row_names = FALSE, top_annotation = ha, row_km = 5) +
Heatmap(base_mean, name = "base mean")
```
The one column heatmap at the side represent the mean value of the gene expression value. Now if we would like to and mean value of all five clusters and add to the top of the mean expression heatmap. 

```{r}
ha_mean = HeatmapAnnotation(summary = anno_summary(gp = gpar(fill = 2:6), height = unit(2, "cm")))
Heatmap(mat_scaled, name = "Expression", col = colors, row_title = NULL, show_row_dend = FALSE, show_row_names = FALSE, top_annotation = ha, row_km = 5) +
Heatmap(base_mean, name = "base mean", top_annotation = ha_mean)
```
We have type of genes column in our gene expression matrix, which we can use to make another heatmap beside the base mean heatmap to add gene type information for every row. 
```{r}
Heatmap(mat_scaled, name = "Expression", col = colors, row_title = NULL, show_row_dend = FALSE, show_row_names = FALSE, top_annotation = ha, row_km = 5) +
Heatmap(base_mean, name = "base mean", top_annotation = ha_mean) +
Heatmap(expr$type, name = "gene type")
```



















